---
title: "Class 8 (Solution)"
format: html
editor: visual
---

# Implementing a Bottleneck Neural Network in R

## Introduction

Welcome to our exploration of Bottleneck Neural Networks (BNNs) using the `keras` library in R. Today, we will delve into how BNNs can be a powerful tool for dimensionality reduction, especially in handling complex datasets like our `SportsCars` dataset. By the end of this class, you will have a hands-on understanding of building, training, and evaluating a BNN model in R.

**Objective**: Gain practical experience in implementing a BNN and understand its applications in data analysis.

**Real-World Application**: We'll explore how BNNs are revolutionizing the way we approach high-dimensional data in fields like automotive design, consumer behavior analysis, and more.

## Overview of Bottleneck NNs

Neural Networks Basics: Explain the basics and importance of neural networks.

What is a Bottleneck Neural Network?: Describe the BNN and its role in dimensionality reduction.

## Data Preprocessing

```{r}

```

## Building the BNN Model

```{r}

```

## Training the BNN Model

Hyperparameters Explanation: Explain epochs, batch size, and verbosity.

```{r}

```

## Reconstruction Error

```{r}

```

## Visualizing Reduced Dimensions

```{r}


```
