---
title: "Lab 7"
format: docx
editor: visual
toc: TRUE
---

# Lab 7

This notebook will guide you through the process of building a neural network model using keras and tensorflow in R. We will be working with a motor third party liability (MTPL) insurance dataset.

Load the following libraries

```{r}
library(keras)
library(tensorflow)
library(dplyr)
library(readr)
```

However, you will need to install python on your local machine as keras and tensorflow act only as interfaces. You can do this by following the steps on Canvas from Week 4 Material "Installation guide_python_reticulate_keras.txt". I will also upload this guide on github. Essentially, you need to run this line of code:

```{r}
# Sys.setenv(
#   RETICULATE_AUTOCONFIGURE='FALSE',
#   RETICULATE_MINICONDA_PATH="C:/F71RA-Miniconda/",mustWork = FALSE
# )
# source(system.file("helpers", "install.R", package = "ISLR2"))
# install_miniconda_and_tensorflow()

```

**Please add comments to the codes below outlining what you think each line does!**

## Data Preprocessing

First, we will load and preprocess the data. Preprocessing includes converting character variables to factors, capping certain variables, and splitting the data into a learning set and a test set.

```{r}
# Load your dataset here
dat <- read_csv("freMTPL2freq.csv")


for(i in seq_along(dat)){
  if(is.character(dat[[i]])){
    dat[[i]] <- factor(dat[[i]])
  }
}

dat$ClaimNb <- pmin(dat$ClaimNb, 4)
dat$Exposure <- pmin(dat$Exposure, 1)

## Learn Test Split 

learn_idx <- sample(1:nrow(dat), round(0.9*nrow(dat)), replace = FALSE) # learning sample index
learn <- dat[learn_idx, ]
test <- dat[-learn_idx, ]

n_l <- nrow(learn)
n_t <- nrow(test)
```

## Data Preparation for Neural Network

Add comments to the codes below outlining what you think each line does.

```{r}
#To do: crate the following matrices
# - Design Matrix for continuous variables
# - Matrices for the categorical variables
# - Matrix for offset 

MM_scaling <- function(x){ 2*(x-min(x))/(max(x)-min(x)) - 1}

dat_NN <- data.frame(ClaimNb = dat$ClaimNb)
dat_NN$DriveAge <- MM_scaling(dat$DrivAge)
dat_NN$BonusMalus <- MM_scaling(dat$BonusMalus)
dat_NN$Area <- MM_scaling(as.integer(dat$Area))
dat_NN$VehPower <- MM_scaling(as.numeric(dat$VehPower))
# dat_NN$VehAge <- MM_scaling(as.numeric(dat$VehAgeGLM))
dat_NN$VehAge <- MM_scaling(as.numeric(dat$VehAge))
dat_NN$Density <- MM_scaling(dat$Density)
dat_NN$VehGas <- MM_scaling(as.integer(dat$VehGas))

## Learn Test Split 

learn_NN <- dat_NN[learn_idx,]
test_NN <- dat_NN[-learn_idx,]

## 

Design_learn <- as.matrix(learn_NN[ , -1])
Design_test <- as.matrix(test_NN[ ,-1])

# DrivAge_learn <- as.matrix(learn_NN$DriveAge)
# DrivAge_test  <- as.matrix(test_NN$DriveAge)
# BonusMalus_learn
# BonusMalus_test

Br_learn <- as.matrix(as.integer(learn$VehBrand)) - 1 
Br_test <- as.matrix(as.integer(test$VehBrand)) - 1

Re_learn <- as.matrix(as.integer(learn$Region)) - 1
Re_test <- as.matrix(as.integer(test$Region)) - 1


Vol_learn <- as.matrix(learn$Exposure)
Vol_test <- as.matrix(test$Exposure)
LogVol_learn <- log(Vol_learn)
LogVol_test <- log(Vol_test)

Y_learn <- as.matrix(learn_NN$ClaimNb)
Y_test <- as.matrix(test_NN$ClaimNb)

```

## Neural Network Architecture

We will define a neural network architecture with embedded layers for categorical predictors and several dense layers for the main architecture.

```{r}
# ------------------ Step 2: Neural Network -------------------------------------------------------

# ------------------ Step 2.1: Hyperparameters initialization -------------------------------------

q1 <- 20 # the dimension of the 1st hidden layer
q2 <- 15 # the dimension of the 2nd hidden layer
q3 <- 10 # the dimension of the 3rd hidden layer
qEmb <- 2 # the dimension of the embedded layer for "VehBrand" and "Region"

epochs <- 200 # number of epochs to train the model
batchsize <- 10000 # number of samples per gradient update

# ------------------ Step 2.2: Input layer --------------------------------------------------------
  
  # Input layer for continuous features
# DesignShape <- ncol(learn_NN) - 1 # the number of the continuous predictors

Design <- layer_input(shape = ncol(learn_NN) - 1, dtype = 'float32', name = 'Design')

# Input layer for categorical features

Br_ndistinct <- length(unique(learn$VehBrand)) # number of vehicle brands = 11
Re_ndistinct <- length(unique(learn$Region)) # number of regions = 21

VehBrand <- layer_input(shape = 1, dtype = 'int32', name = 'VehBrand')
Region <- layer_input(shape = 1, dtype = 'int32', name = 'Region')

# Input layer for Exposure (as the offset)
LogVol <- layer_input(shape = 1, dtype = 'float32', name = 'LogVol')
Vol <- layer_input(shape = 1, dtype = 'float32', name = 'Vol')

# ------------------ Step 2.3: Embedding layer ----------------------------------------------------

# Create embedding layer for categorical predictors (dimension: qEmb = 2)
# Vehicle Brand: 11 -> 2
# Region: 21 -> 2

BrEmb = VehBrand %>%
  layer_embedding(input_dim = Br_ndistinct, output_dim = qEmb, input_length = 1, name = 'BrEmb') %>%
  layer_flatten(name = 'Br_flat')
ReEmb = Region %>%
  layer_embedding(input_dim = Re_ndistinct, output_dim = qEmb, input_length = 1, name = 'ReEmb') %>%
  layer_flatten(name = 'Re_flat')


# ------------------ Step 2.4: Main architecture and Output layer ---------------------------------

# Main architecture with 3 hidden layers
Network <- list(Design, BrEmb, ReEmb) %>% layer_concatenate(name = 'concate') %>%
  
  # 1st hidden layer
  layer_dense(units = q1, activation = 'tanh', name = 'hidden1') %>%
  
  # 2nd hidden layer
  layer_dense(units = q2, activation = 'tanh', name = 'hidden2') %>%
  
  # 3rd hidden layer
  layer_dense(units = q3, activation = 'tanh', name = 'hidden3') %>%
  
  # provide one neuron in the output layer
  layer_dense(units = 1, activation = 'linear', name = 'Network')


# Output layer to combine the main architecture and the offset layer (Exposure)
Response = (Network + LogVol) %>%
  
  # give the response
  layer_dense(units = 1,
              activation = 'exponential',
              name = 'Response',
              trainable = FALSE,
              weights = list(array(1, dim = c(1,1)), array(0, dim = c(1))))

# ------------------ Step 2.5: Model configuration and fitting ------------------------------------

# Model assembly
model <- keras_model(inputs = c(Design, VehBrand, Region, LogVol), outputs = c(Response))

summary(model)

# Model configuration
model %>% compile(
  loss = 'poisson', # set poisson deviance loss function as the objective loss function
  optimizer = 'nadam'
)

# Model fitting by running gradient descent method to minimize the objective loss function
{ 
  
  t1 <- proc.time()
  
  fit <- model %>% fit(
    
    list(Design_learn, Br_learn, Re_learn, LogVol_learn), # all predictors
    Y_learn, # response
    
    verbose = 1, # verbose = 0 silences the progress bar for the process
    # verbose = 1 shows the fitting process, incl. learning loss and validation loss, epoch by epoch
    
    epochs = epochs, # epochs = 1,000
    
    batch_size = batchsize, # batchsize = 10,000
    
    validation_split = 0.2 # 20% as validation set
    
  )
  
  print(proc.time()-t1)
}


```

## Results

```{r}
# Predicted value of the claim numbers
learn$nn0 <- as.vector(model %>% predict(list(Design_learn, Br_learn, Re_learn, LogVol_learn)))
test$nn0 <- as.vector(model %>% predict(list(Design_test, Br_test, Re_test, LogVol_test)))

dev.loss <- function(y, mu, density.func) {
  logL.tilde <- log(density.func(y, y))
  logL.hat <- log(density.func(y, mu))
  2 * mean(logL.tilde - logL.hat)
}

dev.loss(y = learn$ClaimNb, mu = learn$nn0, dpois)
```

### Task: try and change the loss function from poisson to negative binomial.