---
title: "Lab 1: R Foundations for Risk and Insurance"
subtitle: "F71RA Machine Learning for Risk and Insurance 1"
format: docx
editor: visual
toc: true
---

## Before you start

**Learning objectives.** By the end of this lab you should be able to

-   create and manipulate matrices in R (multiplication, inversion, transposition) and recognise when an operation is not defined;
-   compute and interpret summary statistics for insurance data, with particular attention to skewness and heavy tails;
-   simulate data from a known model, fit a linear regression with `lm()`, and check whether the fitted model recovers the truth.

**Packages.** Base R covers most of this lab. For Exercise 2.5 you will need the `moments` package:

```{r}
#| eval: false
install.packages("moments")
```

**Reproducibility.** Several exercises use random numbers. Run the chunk below once at the start of your session so that your "random" values match the model answers.

```{r}
set.seed(71)
```

**Submission.** Complete every chunk, render the document, and submit the rendered file. Where an exercise asks a question in words, answer it in a sentence or two directly below the chunk.

------------------------------------------------------------------------

## 1. Matrix algebra

Almost every method in this course, from linear regression to neural networks, is a sequence of matrix operations. These exercises make sure you are comfortable with the basic building blocks in R.

### Exercise 1.1: Creating a matrix

Create a $3 \times 3$ matrix $M$ whose entries are random integers between 1 and 10 (use `sample()`). Print it, then print `dim(M)`.

```{r}

```

Now create the same matrix with `byrow = TRUE`. In one sentence, explain what changed and why.

```{r}

```

### Exercise 1.2: Matrix multiplication

Generate a random $4 \times 3$ matrix $A$ and a random $3 \times 2$ matrix $B$, and compute the product $C = AB$ using `%*%`.

```{r}

```

Before you run any code: what are the dimensions of $C$? Check your answer with `dim(C)`. Then try `B %*% A` and explain the error message. Finally, compare `A %*% B` with `A * B` (or explain why the second one fails).

```{r}

```

### Exercise 1.3: Matrix inversion

Not every square matrix has an inverse. Create the following two matrices and compute `det()` for each:

$$
X_1 = \begin{pmatrix} 2 & 1 \\ 1 & 3 \end{pmatrix}, \qquad
X_2 = \begin{pmatrix} 2 & 4 \\ 1 & 2 \end{pmatrix}.
$$

```{r}

```

Use `solve()` to invert $X_1$, and verify that $X_1 X_1^{-1}$ is the identity matrix. Then try to invert $X_2$. What happens, and how could you have predicted it from the determinant?

```{r}

```

### Exercise 1.4: Transposition

Generate a random $4 \times 2$ matrix $D$ and compute its transpose with `t()`. Then compute $D^\top D$ and $D D^\top$. What are their dimensions, and which one is more likely to be invertible? (You will meet $D^\top D$ again in Section 3.)

```{r}

```

------------------------------------------------------------------------

## 2. Descriptive statistics for claims data

In this section we work with a small simulated motor insurance portfolio. Claim sizes in practice are strongly right-skewed: most claims are small, and a few are very large. We simulate this with a lognormal distribution.

### Exercise 2.1: Building the dataset

Run the chunk below to create a data frame `claims` with 500 policies. Each row records the driver's age, the vehicle's age, the sum insured, and the size of one claim.

```{r}
n <- 500
claims <- data.frame(
  driver_age  = sample(18:80, n, replace = TRUE),
  vehicle_age = sample(0:15, n, replace = TRUE),
  sum_insured = round(runif(n, 5000, 40000), -2)
)
claims$claim_size <- exp(7 + 0.00003 * claims$sum_insured
                         - 0.01 * claims$driver_age + rnorm(n, 0, 0.8))
head(claims)
```

Use `str()` and `summary()` to inspect the data. Which variable has the largest gap between its mean and its median? What does that tell you about its distribution?

```{r}

```

### Exercise 2.2: Variance and standard deviation

Compute the variance and standard deviation of `claim_size` using `var()` and `sd()`.

```{r}

```

R's `var()` divides by $n-1$, not $n$. Compute the population variance by hand (i.e. $\frac{1}{n}\sum (x_i - \bar{x})^2$) and confirm the two differ by a factor of $(n-1)/n$. Why does R use $n-1$ by default?

```{r}

```

### Exercise 2.3: Histograms

Draw a histogram of `claim_size` with roughly 30 bins. Then draw a histogram of `log(claim_size)`. Describe the shape of each in one sentence.

```{r}

```

Why do actuaries so often model claim sizes on the log scale? What would go wrong if you fitted a normal distribution directly to `claim_size`?

### Exercise 2.4: Correlation

Compute the correlation matrix of `driver_age`, `vehicle_age`, `sum_insured` and `claim_size` using `cor()`.

```{r}

```

Which variables are most strongly related to `claim_size`? Do the signs match how the data were generated in Exercise 2.1? Produce a scatter plot of `claim_size` against the strongest predictor.

```{r}

```

### Exercise 2.5: Skewness and kurtosis

Skewness measures asymmetry; a right-skewed distribution has a long right tail and positive skewness. Kurtosis measures how heavy the tails are relative to a normal distribution, which has kurtosis 3.

Using the `moments` package, compute the skewness and kurtosis of `claim_size` and of `log(claim_size)`.

```{r}
library(moments)

```

Now simulate 500 values from $N(0, 1)$ and compute the same two quantities. Comment on the differences. Why do heavy tails matter for an insurer more than for, say, a retailer analysing customer heights?

```{r}

```

------------------------------------------------------------------------

## 3. Linear regression

We now fit our first predictive model. The key idea in this section is to **simulate data from a model whose parameters you know**, fit the model, and check whether the fit recovers the truth. This is a habit you should keep for every method in the course.

### Exercise 3.1: Simple linear regression

Simulate 200 observations from the model

$$
Y = 2 + 3X + \varepsilon, \qquad X \sim N(0,1), \quad \varepsilon \sim N(0, 1).
$$

Store them in a data frame `sim`.

```{r}

```

Fit the model `lm(Y ~ X, data = sim)`, print `coef()`, and plot the data with the fitted line added using `abline()`. How close are the estimated intercept and slope to 2 and 3?

```{r}

```

### Exercise 3.2: Multiple linear regression

Add a second predictor $Z \sim N(0,1)$ and regenerate $Y$ from

$$
Y = 2 + 3X - 1.5Z + \varepsilon .
$$

Fit `lm(Y ~ X + Z, data = sim)` and look at `summary()`.

```{r}

```

Answer the following in words:

1.  What does the coefficient on $Z$ mean in plain language?
2.  Is each coefficient statistically significant? Which column of the summary tells you this?
3.  What does the $R^2$ value tell you, and why is it not 1 even though you know the true model?

### Exercise 3.3: Regression on real-ish data

Return to the `claims` data from Section 2. Fit a linear regression of `log(claim_size)` on `driver_age`, `vehicle_age` and `sum_insured`.

```{r}

```

Compare the fitted coefficients with the values used to generate the data in Exercise 2.1. Which predictor is doing nothing, and does the model agree?

------------------------------------------------------------------------

## Stretch exercise: regression by hand

The `lm()` function solves the *normal equations*. For a design matrix $\mathbf{X}$ (with a column of ones for the intercept) and response vector $\mathbf{y}$, the least-squares estimate is

$$
\hat{\boldsymbol\beta} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{y}.
$$

Using only the matrix tools from Section 1 (`cbind()`, `t()`, `%*%`, `solve()`), compute $\hat{\boldsymbol\beta}$ for the model in Exercise 3.2 and confirm it matches `coef()` from `lm()`.

```{r}

```

Why did we need $\mathbf{X}^\top \mathbf{X}$ to be invertible? What would happen if $Z$ were an exact linear function of $X$? Try it.
