---
title: "Lab 1"
format: docx
editor: visual
toc: true
---

## 1: Matrix Algebra Exercises

### Exercise 1.1: Matrix Creation

Creating a matrix involves arranging numbers in rows and columns. Create a $3\times3$ matrix $M$ with random integer values ranging from 1 to 10.

```{r}

```

### Exercise 1.2: Matrix Multiplication

Matrix multiplication is a fundamental operation in linear algebra. Generate two random matrices $A$ $(4\times3)$ and $B$ $(3\times2)$ and compute their product $C$.

```{r}

```

### Exercise 1.3: Matrix Inversion

Matrix inversion is the process of finding the ”inverse” of a matrix, denoted as $A^{−1}$. Create a $2x2$ matrix `X` and calculate its inverse `X_inv` if it exists.

```{r}

```

### Exercise 1.4: Matrix Transposition

Matrix transposition involves switching the rows and columns of a matrix. Generate a random $4\times2$ matrix $D$ and calculate its transpose $D^\top$ using the `t()` function.

```{r}

```

## 2: Descriptive and Summary Statistics Exercises

### Exercise 2.1: Data Creation

Generate a dataset with 100 random values for $X\sim N(0,1)$.

```{r}

```

### Exercise 2.2: Variance and Standard Deviation

Variance and standard deviation are measures of the spread or dispersion of a dataset. Calculate the variance and standard deviation of the $X$ variable.

```{r}

```

### Exercise 2.3: Histogram

A histogram is a graphical representation of the distribution of data. Create a histogram for the $X$ variable with 10 bins using the `hist()` function.

```{r}


```

### Exercise 2.4: Correlation Matrix

Generate a dataset with two variables $X$ and $Y$ and calculate their correlation matrix.

```{r}


```

### Exercise 2.5: Skewness and Kurtosis

Skewness is a measure of symmetry, or more precisely, the lack of symmetry. A distribution, or data set, is symmetric if it looks the same to the left and right of the center point. Kurtosis is a measure of whether the data are heavy-tailed or light-tailed relative to a normal distribution. Compute the skewness and kurtosis of the $X$ variable.

```{r}

```

## 3: Linear Regression Exercises

### Exercise 3.1: Data Generation

Generating a synthetic dataset involves creating a dataset with known properties for testing purposes. Create a synthetic dataset with two variables $X$ and $Y$ for linear regression analysis.

```{r}


```

### Exercise 3.2: Multiple Linear Regression

Multiple linear regression involves using multiple predictors to predict a response variable. Extend the dataset to include another predictor $Z$ and fit a multiple linear regression model using the `lm()` function. Now try and interpret this model using the `summary()` function.

```{r}

```
