---
title: "Class 7 (Solution)"
format: html
editor: visual
---

# Class 7 (5th Nov 2024)

This notebook will guide you through the process of building a neural network model using keras and tensorflow in R. We will be working with a motor third party liability (MTPL) insurance dataset.

Load the following libraries

```{r}


```

However, you will need to install python on your local machine as keras and tensorflow act only as interfaces. You can do this by following the steps on Canvas from Week 4 Material "Installation guide_python_reticulate_keras.txt". I will also upload this guide on github. Essentially, you need to run this line of code:

```{r}


```

**Please add comments to the codes below outlining what you think each line does!**

## Data Preprocessing

First, we will load and preprocess the data. Preprocessing includes converting character variables to factors, capping certain variables, and splitting the data into a learning set and a test set.

```{r}

```

## Data Preparation for Neural Network

Add comments to the codes below outlining what you think each line does.

```{r}

```

## Neural Network Architecture

We will define a neural network architecture with embedded layers for categorical predictors and several dense layers for the main architecture.

```{r}

```

## Results

```{r}

```
